Exploring probability distributions for bivariate temporal granularities

### 2019-11-03 main.Rmd
revising section headers

<!-- general calendar categorization -->
Often we partition time into months, weeks, days and so on in a hierarchical manner to relate it to data. Such discrete abstractions of time can be thought of as time granularities [@aigner2011visualization]. Examples of time abstractions may also include day-of-week, time-of-day, week-of-year, day-of-month, month-of-year, working day/non-working day, etc which are useful to represent different periodicities in the data. Let us call all of these different abstractions of time as "calendar categorizations".

These calendar categorizations can be linear, circular or aperiodic. The calendar categorizations as "linear" if they respect the linear progression of time. We call these **linear time granularities**. Examples include hours, days, weeks and months. **Circular time granularities** accommodate periodicities in time such as hour of the day, and day of the week. **Aperiodic time granularities** are neither linear nor circular, such as day of the month or public holidays.

@@

## Linear {#linear-gran-def}

@@

## Circular {#circular-gran-def}

Periodicity is very common in all kinds of data and circular granularities can be construed to reflect such periodicities. Circular granularities can be constructed by putting to use two linear granularities. However, mappings between linear granularities can be regular or irregular. For example, a regular mapping exists between minutes and hours, where 60 minutes always add up to 1 hour. On contrary, an irregular mapping exists between days and months, since one might can have days ranging from 28 to 31. Hence, circular granularities can also be "periodic" or "aperiodic" depending on

@@

## Aperiodic {#aperiodic-gran-def}
<!-- Formal definition of aperiodic circular granularity -->

@@

Identifying repeating (periodic/aperiodic) patterns are necessary in revealing patterns and future trends of a temporal data. Often there is a need for periodicity detection to find whether and how frequent a periodic/aperiodic pattern is repeated within the series. To consider the exhaustive set of temporal regularities that might exist in the data, we can categorize the set of temporal granularities to single or multiple order up granularities.

## Definitions of time granularities based on order

@@

 -  The first column represents the (temporal) units in ascending order of hierarchy. Any two units can join together to form a granularity.

@@

### Single-order-up granularities

@@

### Computation of multiple order-up granularities

@@

#### All single-order-up granularities are periodic

@@

#### Mixed single-order-up granularities - circular or aperiodic

@@

# Interaction

@@

\autoref{fig:allFig} (a) shows the letter value plot of electricity consumption of Victoria across days of the month for few months like January, February, April and December. Letter value plots convey detailed information about tails of the distribution and outliers are unexpected observations rather than extreme observations. M, F, E, D and C represents 50%, 25%, 12.5%, 6.25% and 3.13% of the tail area respectively. Some observations that can be made from these letter-value plots include a) Right tails for most days in January lying above the median is more extended than the left tails contrary to days in April which has longer left tails. b) Days in mid January and February are characterized by high variation in consumption level, but days in January have longer right tails. c) Last five days in December (Christmas holidays) shows low variation in consumption with longer left tails implying people typically consume less electricity. We can conclude that Month-of-Year and Day-of-Month are harmonies.

\autoref{fig:allFig} (c) shows box plot of electricity consumption of Victoria from 2012 to 2014 across days of the year by the 1st, 15th, 29th and 31st days of the month. The box plot is a a very compact distributional summary, displaying median, quartile boundaries, hinges, whiskers and outliers. All facets do not contain data across same x-axis levels leading to difficulty in comparison across facets. Hence, Day-of-Month and Day-of-Year are clashes.

Take another example \autoref{fig:allFig} (d) showing violin plot across days of the month faceted by week of the month. Violin plots are a combination of box plot and density plot. Here, the first week of the month correspond to certain days of the month which are different for different week of the month. These kinds of graphics can hinder our comprehension across different weeks of the month and can be categorized as a incompatible combination to plot.

In \autoref{fig:allFig} (e), variations across week of the year conditional on week of the month can be observed through a ridge plot. The y-axis represents week of the year and the x-axis represents electricity consumption. Ridge plots are density plots all aligned to the same horizontal scale and presented with a slight overlap. They are designed to bring out changes in distributions over time. The data is distributed unequally for different levels of weeks of the year for different facets, which hinders comparison across facets. So, Week-of-Month and Week-of-Year are clashes.

\autoref{fig:allFig} (f) shows decile plots of consumption across  day of the year and month of the year. Decile plots are useful in displaying distribution through deciles without much clutter. This plot, however, seem ineffective for comparison as levels of x-axis for which observed data is plotted are completely disjoint across facets. So Month-of-Year and Day-of-Year are clashes as well.

From the above examples, we can see that the choices of the circular granularities are harmonies while some are clashes. Also since some choices work and others don't, it must be the attributes of the circular granularities or their relationships which are in play in deciding if the resulting plot would be a good candidate for exploratory analysis.

@@

VIC %>% 
  create_gran("month_year") %>% 
  filter(month_year %in% c("Jan", "Mar", "July", "Dec")) %>% 
  prob_plot("month_year","day_month",
            response = "Demand",
            plot_type = "lv") + 
  ylab("Electricity Demand [KWh]") +
  ggtitle("(a) Letter value plot")

@@

When we plot some observations with two circular or aperiodic granularities used as aesthetics or facets, problems might arise when we have empty combinations. It is important to check how these two granularities interact with each other, before we expect to gain insights by plotting them together.

For example, suppose we have two circular or aperiodic granularities $C_1$ and $C_2$, such that $C_1$ maps row numbers to a set $\{A,B,C,D$\}, and $C_2$ maps row numbers to a set $\{X,Y,Z\}$.  That is, let $S_{ij}$ be the set of row numbers such that for all $s \in S_{ij}$, $C_1(s) = i$ and $C_2(s) = j$. In this example, we have 12 such sets $S_{ij}$ because $i$ can take 4 values and $j$ can take 3 values. The graphs that don't work are those where many of these 12 sets are empty. Now, many situations can lead to any of these sets being empty. Let us discuss the following cases, where one or more of these sets can be empty.

@@

The combinations of circular/aperiodic granularities which lead to structurally, event-driven or build-based empty-combinations are referred to as to as **clashes**.
And the ones that ones that do not lead to any missing combinations are called **harmonies** as they promote the exploratory analysis through visualization.

@@

In \autoref{fig:allFig} (c), we have empty combinations when we plot observations with "Day-of-Month" and "Day-of-Year". Here, the 1st day of the month can never correspond to 2nd, 3rd or 4th day of the year. Hence,  "Day-of-Month" and "Day-of-Year" are clashes due to the way they map to calendar.

In \autoref{fig:allFig} (a), we will not have any empty combinations because every DoM can occur in all MoY. Here, mapping from days to months is irregular in the sense that one month can consist of 28, 29, 30 or 31 days. There is no denying that the 29th, 30th or 31st day of the month needs to be analysed with caution due to the irregular mapping, however, the first 28 days of the month will occur in all months of the year.

More generally, if we have two circular granularities C1 and C2 which has [A, B, C, D] and [X, Y, Z] categories/levels. When C1 and C2 are used as aesthetics or facets, problems arise when we encounter empty combinations. Let $S_{i,j}$ be the set of combination of the levels of C1 and C2. In this example, we have 12 such sets $S_{i,j}$ because i can take 4 values and j can take 3 values. The graphs that don't work are those where many of these 12 sets are empty. In other words, if there are levels of the x-axis which are not spanned by levels of the faceted variable or vice versa we will have structurally empty sets leading to potential ineffective graphs.

Thus, the common link that differentiates harmonies from clashes are:

  a) There should not be any levels of the faceted variable which is empty for one or more levels of the factor plotted across x-axis.

  b) There should not be any level of the factor plotted across x-axis which doesn't have values for all levels of factors plotted across facets.

@@

## Advice algorithm for exploring conditional probability distributions

@@

@@

@@

<!-- ```{r read_data, echo=FALSE} -->

@@

<!-- cricket_tsibble_all %>% -->
<!--   granplot( -->
<!--     gran1 = "inning_match", -->
<!--     gran2 = "over_inning", -->
<!--     hierarchy_model, -->
<!--     response = "total_runs", -->
<!--     plot_type = "quantile" -->
<!--   ) -->
<!-- ``` -->

>>>>

<!-- general calendar categorization -->
Often we partition time into months, weeks or days to relate it to data. Such discrete abstractions of time can be thought of as time granularities [@aigner2011visualization]. Examples of time abstractions may also include day-of-week, time-of-day, week-of-year, day-of-month, month-of-year, working day/non-working day, etc which are useful to represent different periodicities in the data.

## Arrangement: Linear vs. Circular vs. aperiodic

The arrangement of the time domain can result in deconstructing time in linear, circular or nearly circular ways. Time granularities are **linear** if they respect the linear progression of time. Examples include hours, days, weeks and months. **Circular** time granularities accommodate periodicities in time such as hour of the day, and day of the week. **Aperiodic** time granularities are neither linear nor circular, such as day of the month or public holidays.

@@

### Linear {#linear-gran-def}

@@

### Circular {#circular-gran-def}

Periodicity is very common in all kinds of data. Circular granularities, which are constructed by using two linear granularities, can be utilized to explain such periodicities. However, mappings between linear granularities can be regular or irregular. For example, a regular mapping exists between minutes and hours, where 60 minutes always add up to 1 hour. On contrary, an irregular mapping exists between days and months, since one might can have days ranging from 28 to 31. Hence, circular granularities can also be "periodic" or "aperiodic" depending on if the mapping is regular or irregular.

@@

### Aperiodic {#aperiodic-gran-def}
<!-- Formal definition of aperiodic circular granularity -->

@@

<!-- Identifying repeating (periodic/aperiodic) patterns are necessary in revealing patterns and future trends of a temporal data. Often there is a need for periodicity detection to find whether and how frequent a periodic/aperiodic pattern is repeated within the series. -->

To consider the exhaustive set of temporal regularities that might exist in the data, we can also define temporal granularities based on the hierarchical structure of a calendar.

## Order: Single vs. Multiple

@@

 -  The first column represents the (temporal) units in ascending order of hierarchy. Any two units, which are essentially the linear granularities, can join together to form a circular/aperiodic granularity.

@@

Computation of time granularities would differ in a periodic and non-periodic set up.

## Single-order-up granularities

@@

## Multiple order-up granularities

@@

### All single-order-up granularities are periodic

@@

### Mixed single-order-up granularities - circular or aperiodic

@@

# Interaction of time granularities

@@

<!-- What  is interaction and some examples why it is important to study interaction -->

Before exploring the behavior of a "dependent variable" across time granularities,
it is important to know how these granularities interact with each other. If two time granulaties of interest interact, the relationship between each of the interacting variables and the dependent variable will depend on the value of the other interacting variable, For illustration, distribution of half-hourly electricity consumption of Victoria is plotted across different time granularities in each of the panel in
\autoref{fig:allFig}. \autoref{fig:allFig} (a) shows the letter value plot across days of the month faceted by months like January, March and December. \autoref{fig:allFig} (c) shows box plot  across days of the year by the 1st, 15th, 29th and 31st days of the month. \autoref{fig:allFig} (d) showing violin plot across days of the month faceted by week of the month. \autoref{fig:allFig} (e), variations across week of the year conditional on week of the month can be observed through a ridge plot and \autoref{fig:allFig} (f) shows decile plots across  day of the year and month of the year.

Clearly, in \autoref{fig:allFig}, we observe that some choices of time granularities work and others do not. In \autoref{fig:allFig} (c), there will be no observations for some combinations "Day-of-Month" and "Day-of-Year". In particular, the 1st day of the month can never correspond to 2nd, 3rd or 4th day of the year. On the contrary, for \autoref{fig:allFig} (a), we will not have any combinations with zero combinations because every "Day-of-Week" can occur in any "Month-of-Year". Thus the graphs that don't work are those where many of the combination sets are empty. In other words, if there are levels of time granularities plotted across x-axis which are not spanned by levels of the time granularities plotted across facets or vice versa, we will have empty sets leading to potential ineffective graphs. We hypothesize that the interaction of these  granularities are in play while deciding if the resulting plot would be a good candidate for exploratory analysis.

<!-- \autoref{fig:allFig} (c) shows box plot of electricity consumption of Victoria from 2012 to 2014 across days of the year by the 1st, 15th, 29th and 31st days of the month. -->

<!-- Letter value plots convey detailed information about tails of the distribution and outliers are unexpected observations rather than extreme observations. M, F, E, D and C represents 50%, 25%, 12.5%, 6.25% and 3.13% of the tail area respectively. Some observations that can be made from these letter-value plots include a) Right tails for most days in January lying above the median is more extended than the left tails contrary to days in April which has longer left tails. b) Days in mid January and February are characterized by high variation in consumption level, but days in January have longer right tails. c) Last five days in December (Christmas holidays) shows low variation in consumption with longer left tails implying people typically consume less electricity. We can conclude that Month-of-Year and Day-of-Month are harmonies. -->

<!-- \autoref{fig:allFig} (c) shows box plot of electricity consumption of Victoria from 2012 to 2014 across days of the year by the 1st, 15th, 29th and 31st days of the month. The box plot is a a very compact distributional summary, displaying median, quartile boundaries, hinges, whiskers and outliers. All facets do not contain data across same x-axis levels leading to difficulty in comparison across facets. Hence, Day-of-Month and Day-of-Year are clashes. -->

<!-- Take another example \autoref{fig:allFig} (d) showing violin plot across days of the month faceted by week of the month. Violin plots are a combination of box plot and density plot. Here, the first week of the month correspond to certain days of the month which are different for different week of the month. These kinds of graphics can hinder our comprehension across different weeks of the month and can be categorized as a incompatible combination to plot. -->

<!-- In \autoref{fig:allFig} (e), variations across week of the year conditional on week of the month can be observed through a ridge plot. The y-axis represents week of the year and the x-axis represents electricity consumption. Ridge plots are density plots all aligned to the same horizontal scale and presented with a slight overlap. They are designed to bring out changes in distributions over time. The data is distributed unequally for different levels of weeks of the year for different facets, which hinders comparison across facets. So, Week-of-Month and Week-of-Year are clashes. -->

<!-- \autoref{fig:allFig} (f) shows decile plots of consumption across  day of the year and month of the year. Decile plots are useful in displaying distribution through deciles without much clutter. This plot, however, seem ineffective for comparison as levels of x-axis for which observed data is plotted are completely disjoint across facets. So Month-of-Year and Day-of-Year are clashes as well. -->

<!-- In \autoref{fig:allFig} (c), we have empty combinations when we plot observations with "Day-of-Month" and "Day-of-Year". Here, the 1st day of the month can never correspond to 2nd, 3rd or 4th day of the year. Hence,  "Day-of-Month" and "Day-of-Year" are clashes due to the way they map to calendar. -->

<!-- In \autoref{fig:allFig} (a), we will not have any empty combinations because every DoM can occur in all MoY. Here, mapping from days to months is irregular in the sense that one month can consist of 28, 29, 30 or 31 days. There is no denying that the 29th, 30th or 31st day of the month needs to be analysed with caution due to the irregular mapping, however, the first 28 days of the month will occur in all months of the year. -->

<!-- From the above examples, we can see that the choices of the circular granularities are harmonies while some are clashes. Also since some choices work and others don't, it must be the attributes of the circular granularities or their relationships which are in play in deciding if the resulting plot would be a good candidate for exploratory analysis. -->

@@

VIC %>% 
  create_gran("month_year") %>% 
  filter(month_year %in% c("Jan", "Mar", "July", "Dec")) %>% 
  prob_plot("month_year","day_week",
            response = "Demand",
            plot_type = "lv") + 
  ylab("Electricity Demand [KWh]") +
  ggtitle("(a) Letter value plot")

@@

To define a general framework, we define harmony and clashes. Harmonies are the pair of granularities that aid the process of exploratory analysis and clashes are pair that interact in a non-compatible ways and hinder the process of exploration of the dependent variable if plotted together.

<!-- When we plot some observations with two circular or aperiodic granularities used as aesthetics or facets, problems arise when we have empty combinations. It is important to check how these two granularities interact with each other, before we expect to gain insights by plotting them together.  -->

Suppose we have two circular or aperiodic granularities $C_1$ and $C_2$, such that $C_1$ maps row numbers to a set $\{A_1,A_2,A_3,\dots, A_n$\}, and $C_2$ maps row numbers to a set $\{B_1,B_2,B_3,\dots, B_m$\}.  That is, let $S_{ij}$ be the set of row numbers such that for all $s \in S_{ij}$, $C_1(s) = i$ and $C_2(s) = j$. Since, $C_1$ has n levels and $C_2$ has m levels there will be $nm$ such sets $S_{ij}$. The graphs that don't work are those where many of these 12 sets are empty. Now, many situations can lead to any of these sets being empty. Let us discuss the following cases, where one or more of these sets can be empty.

@@

Therefore, combinations of circular/aperiodic granularities which lead to structurally, event-driven or build-based empty-combinations are referred to as to as **clashes**.
And the ones that do not lead to any missing combinations are called **harmonies** as they promote the exploratory analysis through visualization.

@@

<!-- More generally, if we have two circular granularities C1 and C2 which has [A, B, C, D] and [X, Y, Z] categories/levels. When C1 and C2 are used as aesthetics or facets, problems arise when we encounter empty combinations. Let $S_{i,j}$ be the set of combination of the levels of C1 and C2. In this example, we have 12 such sets $S_{i,j}$ because i can take 4 values and j can take 3 values.  -->

<!-- Thus, the common link that differentiates harmonies from clashes are: -->

<!--   a) There should not be any levels of the faceted variable which is empty for one or more levels of the factor plotted across x-axis. -->

<!--   b) There should not be any level of the factor plotted across x-axis which doesn't have values for all levels of factors plotted across facets. -->

@@

## Plotting harmonies and clashes

## Homogeneity across and within facets

@@

## Number of levels of the facet variable

@@

```{r}

@@

{r read_data, echo=FALSE}

@@

<!-- cricket_tsibble_all %>% -->
<!--   granplot( -->
<!--     gran1 = "inning_match", -->
<!--     gran2 = "over_inning", -->
<!--     hierarchy_model, -->
<!--     response = "total_runs", -->
<!--     plot_type = "quantile" -->
<!--   ) -->
<!-- ``` --
```
